 

### 2021-09-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is determined by the distribution of tree sizes and tree architecture within a forest, and leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient that are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs. bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In forests with more open upper canopies, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

@@

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table 1). 
Carotenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin (VAZ) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996]. Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defences and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019a; @monsonLeafIsopreneEmission2021].

@@

**Leaf photosynthetic thermal tolerance**
Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020].Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2020]. Thermal tolerance varies globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. Along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotPhotosyntheticPlasticityTropical2021]. Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a]. The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

@@

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs. shade leaves [@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002]. This may vary based on forest type [**REF**], leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015]. Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021].

@@

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, and height, all of which scale with DBH [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
The proportion of GPP allocated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020]. As a result of increasing carbon allocation to these other functions, mass-specific - and sometimes even absolute - woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraCarbonCyclingMature2021].
However, NSC reserves increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

@@

The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments.
While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions [@frittsTreeRingsClimate1976], only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [**Fig. `r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress, ie. resilience [@mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].

There are, however, exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2021].
Potential mechanisms for these patterns include stifling of understory $\lambda E$ under low wind, humid conditions, resulting in above-optimal $T_{leaf}$ (Fig. 3) and lower photosynthesis. Furthermore, warming-induced increases in nighttime respiratory carbon loss may be particularly disadvantageous for understory trees that are operating closer to their light-compensation point than canopy trees. Additional research is required to understand these mechanisms and to disentangle tree growth responses to hot-wet versus hot-dry conditions.

>>>>

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution within a forest is determined by the distribution of tree sizes and tree architecture, along the vertical gradient by leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns that are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs. bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In forests with more open upper canopies, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

@@

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table 1). 
Carotenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin (VAZ) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996]. Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defences and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].

@@

**Leaf photosynthetic thermal tolerance**
Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020]. Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2020]. Thermal tolerance varies globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotPhotosyntheticPlasticityTropical2021]. Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a]. The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

@@

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs. shade leaves [@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002]. This may vary based on forest type [**REF**], acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015], or leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015]. Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021].

@@

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, and height, all of which scale with DBH [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Within a stand, woody growth rates , measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006, Stephenson et al. 2014]. NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

@@

In turn, the climate sensitivity of woody growth tends to be greater in canopy trees. In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [**Fig. `r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [ie. resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
However, exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
Potential mechanisms for these patterns include stifling of understory $\lambda E$ under low wind, humid conditions, resulting in above-optimal $T_{leaf}$ (Fig. 3) and lower photosynthesis. Furthermore, warming-induced increases in nighttime respiratory carbon loss may be particularly disadvantageous for understory trees that are operating closer to their light-compensation point than canopy trees. Additional research is required to understand the mechanisms underlying these intriguing differences, and to disentangle size-related tree growth responses to hot-wet versus hot-dry conditions.
